A Closer Look at SGPV's Small Sample Correction

### 2018-08-15 sgpv_vs_TOST.Rmd
add asymmetric bounds

---
title: "Equivalence Testing and the Second Generation P-Value"
author: "Daniël Lakens & Marie Delacre"
date: "28 juli 2018"
output: github_document
---

@@

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r sgpv_tost, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 1001),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,1000,100), labels = seq(140,150,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
```
*Figure 1*: Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of true population means (x-axis) tested against a mean of 145 in a one-sample *t*-test with a sample size of 1000000 and a standard deviation of 500.

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
```
```{r, include=FALSE}
m <- 145
mu <- 145
sd <- 500
n <- 1000000
low_eqbound = -2 
high_eqbound = 2 
alpha = 0.05

@@

@@

@@

[To be extended]

>>>>

---
title: "Equivalence Testing and the Second Generation P-Value"
author: "Daniël Lakens & Marie Delacre"
date: "15 august 2018"
output: github_document
---

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 1001),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,1000,100), labels = seq(140,150,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 1)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025), col = "grey", lwd = 1, lty = 3)
```
*Figure 1*: Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of true population means (x-axis) tested against a mean of 145 in a one-sample *t*-test with a sample size of 1000000 and a standard deviation of 500.

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 1)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025), col = "grey", lwd = 1, lty = 3)

```
```{r, include=FALSE}
m <- 145
mu <- 145
sd <- 500
n <- 1000000
low_eqbound = -2 
high_eqbound = 2 
alpha = 0.05

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

##The relation between equivalence tests and SGPV when confidence intervals are no symmetrical

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.629 to 0.629, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.992.

The effect of assymetric confidence intervals is most easily noticable at smaller sample sizes, therefore in Figure 11 below we plot the p-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from 1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure 2. The p-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the p-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5. 
```{r, include=FALSE}
n <- 30
r_pop <- .0
low_eqbound_r <- -.45
high_eqbound_r <- .45
alpha <- .05
step <- 0.001

p_tost_list <- numeric(length(seq(-0.99, 0.99, step)))
sgpv_list <- numeric(length(seq(-0.99, 0.99, step)))
p_list <- numeric(length(seq(-0.99, 0.99, step)))
t_list <- numeric(length(seq(-0.99, 0.99, step)))
lowbound_list <- numeric(length(seq(-0.99, 0.99, step)))
highbound_list <- numeric(length(seq(-0.99, 0.99, step)))

for(i in seq(-0.99, 0.99, step)){
  count <- count + 1
  r <- i

  res<-TOSTr(n = n,
             r = r,
             low_eqbound_r = low_eqbound_r,
             high_eqbound_r = high_eqbound_r,
             alpha = alpha,
             plot = FALSE,
             verbose = FALSE)

  t_list[count]=r*sqrt(n-2)/sqrt(1-r^2)
  p_list[count]=2*(1-pt(abs(t_list[count]),df=n-2))
  sgpv_list[count] <- p_delta(res$LL_CI_TTEST,res$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  lowbound_list[count]=res$LL_CI_TTEST
  highbound_list[count]=res$UL_CI_TTEST
} 
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
par(xpd=F)
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, length(seq(-0.99, 0.99, step))),
     yaxt = "n",
     xaxt = "n",
     ylab = "TOST p-value and SGPV",
     xlab = "Correlation")
axis(1, at = seq(0,1981,1981/20), labels = seq(-1,1,0.1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025),
       lty = 2,
       col = "grey")

```
*Figure 11*: Comparison of *p*-values from TOST (black line) and 1-SGPV (dotted grey curve) across a range of observed sample correlations (x-axis) tested against equivalence bounds of r = -0.45 and r = 0.45 with n = 30 and an alpha of 0.05. 
```{r, include=FALSE}
low_eqbound_r = -0.45
high_eqbound_r = 0.45
res<-TOSTr(n = 30,
           r = 0.45,
           low_eqbound_r = -0.45,
           high_eqbound_r = 0.45,
           alpha = .05,
           plot = FALSE,
           verbose = FALSE)
sgpv <- p_delta(res$LL_CI_TTEST,res$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)
```
The reason that the equivalence test and SGPV no longer overlap is also because of asymmetric confidence intervals. If the observed correlation falls exactly on the equivalence bound the *p*-value for the equivalence test indicates that the probability of observing the observed or more extreme data, assuming the equivalence bound is the true effect size, is 50%. In other words, if the true effect size is the same as the equivalence bound, it is equally likely to find an effect more extreme than the equivalence bound, as it is to observe an effect that is less extreme than the equivalence bound. However, as can be seen in Figure 12, the two second generation *p*-values associated with the observed correlations at r = -0.45 and r = 0.45 are `r round(sgpv,3)`. Because the confidence intervals are asymmetric around the observed effect size of 0.45 (ranging from `r round(res$LL_CI_TTEST,2)` to `r round(res$UL_CI_TTEST,2)`) according to Blume et al. (2018) `r round(sgpv,3)*100`% of the data-supported hypotheses are null hypotheses, and therefore `r round(sgpv,3)*100`% of the data-supported hypotheses are compatible with the null premise. 
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
n=30
low_eqbound_r=-.45
high_eqbound_r=.45
r1=-.45
r2=0
r3=.45

res1<-TOSTr(n=n,r=r1,low_eqbound_r=low_eqbound_r,high_eqbound_r=high_eqbound_r,alpha=alpha,plot=FALSE, verbose = FALSE)
res2<-TOSTr(n=n,r=r2,low_eqbound_r=low_eqbound_r,high_eqbound_r=high_eqbound_r,alpha=alpha,plot=FALSE, verbose = FALSE)
res3<-TOSTr(n=n,r=r3,low_eqbound_r=low_eqbound_r,high_eqbound_r=high_eqbound_r,alpha=alpha,plot=FALSE, verbose = FALSE)

LL95_r1=res1$LL_CI_TTEST
UL95_r1=res1$UL_CI_TTEST
LL95_r2=res2$LL_CI_TTEST
UL95_r2=res2$UL_CI_TTEST
LL95_r3=res3$LL_CI_TTEST
UL95_r3=res3$UL_CI_TTEST

par(xpd=F)
plot(NA, ylim=c(0,.7), xlim=c(-1,1), bty="l", yaxt="n", ylab="",xlab="Correlation")
abline(v = c(r1, r2, r3),
       lty = 2,
       col = "grey")

points(x=r1, y=0.6, pch=15, cex=1.1)
segments(LL95_r1,0.6,UL95_r1,0.6, lwd=2)
points(x=r2, y=0.4, pch=15, cex=1.1)
segments(LL95_r2,0.4,UL95_r2,0.4, lwd=2)
points(x=r3, y=0.2, pch=15, cex=1.1)
segments(LL95_r3,0.2,UL95_r3,0.2, lwd=2)
```
*Figure 12*: Three 95% confidence intervals for observed effect sizes of r = -0.45, r = 0, and r = 0.45 for n = 30. Only the confidence interval for r = 0 is symmetric.  
```{r, include=FALSE}
#Example illustrating the most extreme case. 
low_eqbound_r = -0.99
high_eqbound_r = 0.99

res<-TOSTr(n = 4,
           r = 0.99,
           low_eqbound_r = low_eqbound_r,
           high_eqbound_r = high_eqbound_r,
           alpha = .05,
           plot = TRUE,
           verbose = TRUE)
sgpv <- p_delta(res$LL_CI_TTEST,res$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)
sgpv
```

This example illustrates the difference between a proportion and a probability. There is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between two different questions the SGPV and a *p*-value are answers to. The conclusion of this in depth look at asymmetric confidence intervals is that, while a SGPV of 1 or 0 can still be interpreted the same way as a *p*-value of 0.025 and 0.975 can be interpreted in n equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at these values. Although Blume et al (2018, p. 3) state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis'. When the observed correlation equals an equivalence bound, 50% of the normally distributed data is compatible with the null hypothesis (i.e., falls within the equivalence range), even when the SGPV is larger than 0.5.

@@

One weakness of the SGPV is its reliance on the 'small sample correction', where the SGPV is set to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 and the CI overlaps with the upper and lower bounds. This exception to the normal calculation of the SGPV is introduced to prevent misleading values. Without this correction it is possible that a confidence interval is extremely wide, and an equivalence range is extremely narrow, which without the correction would lead to a very low value for the SGPV, which as Blume et al (2018, p. 7) seems to suggest that 'the data favor alternative hypotheses', even when the data is just extremely inaccurate compared to the width of the equivalence range. Although setting the SGPV to 0.5 whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 is necessary to prevent misleading results, it leads to a range of situations where the SGPV is set to 0.5, while the *p*-value from the TOST procedure continues to differentiate (see Figure 6).

As a more extreme example of the peculiar behavior of the 'small sample correction' as currently implemented in the calculation of the SGPV see Figure 13 below. In this figure observed correlations (from a sample size of 10) from -1 to 1 are tested against against an equivalence range from r = 0.4 to r = 0.8. We can see the SGPV has a peculiar shape because it is set to 0.5 for certain observed correlations, even though there is no risk of meaningless SGPV in this range. This example suggests that the current implementation of the 'small sample correction' could be improved.

```{r, include=FALSE}
n <- 10
r_pop <- 0.6
low_eqbound_r <- 0.4
high_eqbound_r <- 0.8
alpha <- .05
step <- 0.001

p_tost_list <- numeric(length(seq(-0.99, 0.99, step)))
sgpv_list <- numeric(length(seq(-0.99, 0.99, step)))
p_list <- numeric(length(seq(-0.99, 0.99, step)))
t_list <- numeric(length(seq(-0.99, 0.99, step)))
lowbound_list <- numeric(length(seq(-0.99, 0.99, step)))
highbound_list <- numeric(length(seq(-0.99, 0.99, step)))

for(i in seq(-0.99, 0.99, step)){
  count <- count + 1
  r <- i

  res<-TOSTr(n = n,
             r = r,
             low_eqbound_r = low_eqbound_r,
             high_eqbound_r = high_eqbound_r,
             alpha = alpha,
             plot = FALSE,
             verbose = FALSE)

  t_list[count]=r*sqrt(n-2)/sqrt(1-r^2)
  p_list[count]=2*(1-pt(abs(t_list[count]),df=n-2))
  sgpv_list[count] <- p_delta(res$LL_CI_TTEST,res$UL_CI_TTEST, low_eqbound_r, high_eqbound_r)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  lowbound_list[count]=res$LL_CI_TTEST
  highbound_list[count]=res$UL_CI_TTEST
} 
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
par(xpd=F)
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, length(seq(-0.99, 0.99, step))),
     yaxt = "n",
     xaxt = "n",
     ylab = "TOST p-values and 1-SGPV",
     xlab = "Correlation")
axis(1, at = seq(0,1981,1981/20), labels = seq(-1,1,0.1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025),
       lty = 2,
       col = "grey")
```
*Figure 13*: Comparison of *p*-values from TOST (black line) and 1-SGPV (dotted grey curve) across a range of observed sample correlations (x-axis) tested against equivalence bounds of r = 0.4 and r = 0.8 with n = 10 and an alpha of 0.05.

### 2018-08-15 sgpv_vs_TOST_small_n_correction.Rmd
add asymmetric bounds

SGPV1 <- TOST_to_SGPV(tost_res = tost_res1)

@@

>>>>

@@

```
Let's look at a single study with diverging p-values to see what is going on.

```{r, include=FALSE}
#
n=40
breaks=c(10,20,40)
d.lo=-0.4
d.hi=0.4

to.stat.1=TOSTone.raw(m=mean(z[1:breaks[1]]), 
                      mu=0,
                      sd=sd(z[1:breaks[1]]),
                      n=length(z[1:breaks[1]]),
                      low_eqbound=d.lo, 
                      high_eqbound=d.hi,
                      alpha=0.05,
                      plot=TRUE)

p.tost.1=max(to.stat.1$TOST_p1,to.stat.1$TOST_p2)

tt.stat.1=t.test(z[1:breaks[1]])$conf.int

p.delta.1=sgpv(est.lo=tt.stat.1[1], est.hi=tt.stat.1[2], 
               null.lo=d.lo, null.hi=d.hi)
p.delta.1
p.delta.2=sgpv(est.lo=to.stat.1$LL_CI_TTEST, est.hi=to.stat.1$UL_CI_TTEST, 
               null.lo=d.lo, null.hi=d.hi)
p.delta.2

p.tost.1
p.delta.1
1-p.delta.1

#Claculate Width of CI
tt.stat.1[2] - tt.stat.1[1]
#Calculate Width of equivalence range
d.hi - d.lo
#Ratio
(tt.stat.1[2] - tt.stat.1[1]) / (d.hi - d.lo)

d.lo=-0.4
d.hi=0.4

to.stat.1=TOSTone.raw(m=0.2, 
                      mu=0,
                      sd=1,
                      n=10,
                      low_eqbound=d.lo, 
                      high_eqbound=d.hi,
                      alpha=0.05,
                      plot=TRUE)

p.tost.1=max(to.stat.1$TOST_p1,to.stat.1$TOST_p2)

p.delta.2=sgpv(est.lo=to.stat.1$LL_CI_TTEST, est.hi=to.stat.1$UL_CI_TTEST, 
               null.lo=d.lo, null.hi=d.hi)
p.tost.1
p.delta.2
